---
title: "MoT Road Safety Data Analysis 1"
author: "Kathlyn Ycong"
date: "13/08/2021"
output: html_document
---
`Project Aim/Goal`
MoT are looking to adapt traficalmr utilities and methodologies for New Zealand 
using open data to understand road safety outcomes

`Project Description`
Adapting UK's trafficalmr to the New Zealand context and enable policy.traffic 
calming intervention analyses.

This project involves data processing with various data such as geospatial 
datasets, APIs and census datasets.

The policy analysis will use the developed data foundations to estimate the 
casualty rate per billion kilometres for walking and cycling during journey 
to work commutes.


`Project Outcomes`
- R functions for processing CAS, census journet to work and, transport relevant 
openstreetmap data using trafficalmr as a guide.
- Estimating causalty rates per billion kilometers for cycling and walking


### Data Inputs

1. Crash Analysis System (CAS) data from NZTA
2. open data from NZTA
3. OpenStreetMap for geospatial data on transport infrastructure
4. Stats NZ census journey to work data


#### Crash data

https://opendata-nzta.opendata.arcgis.com/datasets/8d684f1841fa4dbea6afaefc8a1ba0fc_0

- extracted from the TransportAgency Crash Analysis System(CAS). CAS records all traffic crashes as reported to the Transport Agency by the NZ Police. Not all crashes are reported to the NZ Police. the level of reporting increases with the severity of the crash. Due to the nature of non-fatal crashes it is believed that these are under-reported. CAS covers crashes on all New Zealand roadways or places where the public have legal access with motor vehicle.

#### Open Street Map

https://www.openstreetmap.org/

- A map of the world, createdfor free to use under an open license


#### Stats NZ census journey to work data

- The 2018 Census commuter view dataset contains the employed census usually resident population count aged 15 years and over by statistical area 2 for the main means of travel to work variable from the 2018 census.
The geography corresponds to 2018 boundaries

- This 2018 Census commuter view dataset is displayed by statistical area 2 geography and contains from-to (journey) information on an individual's usual residence and workplace address by main means to travel of work

*Workplace address definition*
- coded from information supplied by respondents about their workplaces. Where respondents do not supply sufficient information, their responses are coded to 'no further defined'. The 2018 Census commuter view datasets excludes these 'not further defined' areas, as such the sum of the counts for each region in this dataset may not be equal to the total employed census usually resident population count aged 15 years and over for that region. 


# Data Analysis


### NZTA Crash data

As noted by Shriv the columns X and Y in CAS data is a GeometryTrype Type. A spatial dataset and the crs is 2193 (usually used 4326). 2193 is a standard specifically for NZ https://epsg.io/2193.

The CAS data consists of the count of object and vehicle type associate to each crash (per row). 

Is there a need to check the interaction other attributes in the data such as weather using regression?

Crash data consists of crashes from 2000 to 2021.

```{r}
## Loading the cast data

#install.packages("iotools")
if (!file.exists("Data/Crash_Analysis_System_(CAS)_data.rds")) {
  cas.data <- iotools::read.csv.raw("Data/Crash_Analysis_System_(CAS)_data.csv")
  saveRDS(cas.data, file="Data/Crash_Analysis_System_(CAS)_data.rds")
} else {cas.data <- readRDS("Data/Crash_Analysis_System_(CAS)_data.rds")}

## link to cas data
#https://opendata-nzta.opendata.arcgis.com/datasets/8d684f1841fa4dbea6afaefc8a1ba0fc_0/explore?location=-9.962041%2C0.000000%2C2.20

#size - 760k rows and 72 columns
dim(cas.data)

#preview the data
#head(cas.data)

#check the data types in the data
#str(cas.data)

#vehicle columns
cas.vehicle <- c('bicycle', 
'bus',
'carStationWagon',
'moped',
'motorcycle',
'otherVehicleType',
'schoolBus',
'suv',
'taxi',
'train',
'truck',
'unknownVehicleType',
'vanOrUtility',
'vehicle')

#Histogram of number of vehicles involve in the crash
#Proportion 1 vechicle crashes involved objects such as tree, cliff and etc.
hist(rowSums(cas.data[cas.vehicle], na.rm = TRUE), main = "Histogram of number of vehicles involve in the crash", ylab = 'no. of crashes', xlab = 'no. of vehicle in the crash')


#Vehicle type crashes
op <- par(mar = c(10,4,2,1) + 0.1)
barplot(colSums(cas.data[cas.vehicle], na.rm = TRUE), las=2, main = '2000 to 2021 Crashes by Vehicle Type')
par(op)
colSums(cas.data[cas.vehicle], na.rm = TRUE)

```


```{r}

serious.injury.per.year <- aggregate(cas.data$seriousInjuryCount, by = list(Crash.Year = cas.data$crashYear), FUN = sum, na.rm = T)

minor.injury.per.year <- aggregate(cas.data$minorInjuryCount, by = list(Crash.Year = cas.data$crashYear), FUN = sum, na.rm = T)

fatality.per.year <- aggregate(cas.data$fatalCount, by = list(Crash.Year = cas.data$crashYear), FUN = sum, na.rm = T)

#plotting the number of crashes
plot(table(cas.data$crashYear), col = 4, ty = 'o', ylab = "number of crashes", main = "The count of crashes per year")

#serious injury crashes
lines(fatality.per.year$Crash.Year, fatality.per.year$x, col = "red", ty = 'o')

#serious injury crashes
lines(serious.injury.per.year$Crash.Year, serious.injury.per.year$x, col = "orange", ty = 'o')

#minor injury crashes
lines(minor.injury.per.year$Crash.Year, minor.injury.per.year$x, col = "purple", ty = 'o')

legend(2008, -10000, legend = c("Total Crashes", "Fatalities", "Serious Injury", "Minor injury"),col=c("blue", "red", "orange", "purple"), bty='n', xpd=NA, horiz=TRUE, xjust=0.5, lwd=c(2,2,2,2))

#average serious injury from crash per year
mean(serious.injury.per.year$x)

#year with the highest crash 
max_val <- max(table(cas.data$crashYear))

#Year with the highest crashes
table(cas.data$crashYear)[table(cas.data$crashYear) == max_val]

```


###STATS NZ Journey to work data
```{r}

#load journey to work data
journey.to.work <-  read.csv("Data/2018-census-main-means-of-travel-to-work-by-statistical-a.csv", fileEncoding="UTF-8-BOM")

#size - 50870    19
dim(journey.to.work)

#columns
# WARNING: noticed column names is different for different PCs - mac vs windows
#names(journey.to.work)

#columns and data type
#str(journey.to.work)

```



```{r}

#proportion of people that works the same area
journey.to.work$same.area.boolean <- journey.to.work[[1]] == journey.to.work[[5]]

same.area.df <-  aggregate(journey.to.work$Total, by=list(same.area=journey.to.work$same.area.boolean), FUN=sum)

barplot(height=same.area.df$x, names=c('different area', 'same area'), col="#69b3a2", main = 'Commuter Workplace vs. Residence Location')

#74% of the commuter work on different area from its residential area
same.area.df$x[1]/sum(same.area.df$x)
same.area.df$x[2]/sum(same.area.df$x)



```

```{r}

#replacing -999 with 0
journey.to.work[journey.to.work == "-999"] <- 0

#most commuters drive a private car/truck/van
op <- par(mar = c(18,4,2,1) + 0.1)
barplot(colSums(journey.to.work[, 9:18]), las = 2)
par(op)

colSums(journey.to.work[, 9:18])
```

### joining statistical area data with journey to work data

```{r}

##loading the data
sa2.centroid.inside <-  read.csv("Data/statistical-area-2-2018-centroid-inside.csv", fileEncoding="UTF-8-BOM")

#names(sa2.centroid.inside)

##joining long and lat data from 'sa2 centroid inside' data with journey to work data

library(dplyr)

#sa2.centroid.inside[c("SA22018_V1_00", "LATITUDE", "LONGITUDE")]

## joining coordinates (geocode) of usual residence 
df <- journey.to.work %>% 
  left_join(sa2.centroid.inside[c("SA22018_V1_00", "LATITUDE", "LONGITUDE")], by = c("SA2_code_usual_residence_address" = "SA22018_V1_00")) %>% 
  rename(usual.residence.LATITUDE = LATITUDE, usual.residence.LONGITUDE = LONGITUDE)

## joining coordinates (geocode) of usual workplace 
df2 <- df %>% 
  left_join(sa2.centroid.inside[c("SA22018_V1_00", "LATITUDE", "LONGITUDE")], by = c("SA2_code_workplace_address" = "SA22018_V1_00")) %>% 
  rename(usual.work.LATITUDE = LATITUDE, usual.work.LONGITUDE = LONGITUDE)

#names(df2)

#dim(df2)


```
### Calculate the distance between two points

```{r}
#install.packages("geosphere")
library(geosphere)

#getting the distance between two points using Haversine formula in meters

df2$dist <- distHaversine(df2[c("usual.residence.LONGITUDE", "usual.residence.LATITUDE")], df2[c("usual.work.LONGITUDE", "usual.work.LATITUDE")])

hist(df2[df2$dist > 0, ]$dist/1e3, breaks=1000, xlim=c(0,50), main = "Histogram of distance travelled by commuter from home to work (km)", xlab = 'distance travelled (km)')

```

### Killed and serious injury by distance travelled

```{r}

#total serious injury and
serious_injury_and_fatal_crashes <- sum(cas.data[cas.data$crashYear %in% 2016:2019, ][c("seriousInjuryCount", "fatalCount")], na.rm = T)

total_dist_travelled <- sum(df2$dist)

#KSI/bkm in crashes between 2016 to 2019
serious_injury_and_fatal_crashes/(total_dist_travelled/1e9)

#KSI/bkm in crashes between 2018
sum(cas.data[cas.data$crashYear == 2018, ][c("seriousInjuryCount", "fatalCount")], na.rm = T)/(total_dist_travelled/1e9)

```
### Using Mapview and osm data to plot traffic calming in Auckland

Note there are different libraries to access openstreet map, either using ggmap, mapview, OpenStreetMap.

```{r}

# this function was from trafficalmr 
#remotes::install_github("saferactive/trafficalmr")

tc_get_osm = function(bbox = NULL, value = NULL, output = "osm_points") {
  res = osmdata::osmdata_sf(
    osmdata::add_osm_feature(
      opq = osmdata::opq(bbox = bbox),
      key = "traffic_calming",
      value = value,
      value_exact = TRUE
    )
  )
  res[[output]]
}


## function used to get traffic calming interventions from OpenStreetMap
traffic_calming_points = tc_get_osm(bbox = "Auckland New Zealand") #tc_get_osm is from osm_data

mapview::mapview(traffic_calming_points["traffic_calming"])


```


### Converting cas data from NZTM projection to geocode (lat/lon)

```{r}

## convert from NZTM projection (EPSG:2193) to lat/lon
## https://www.linz.govt.nz/data/linz-data-service/guides-and-documentation/using-lds-xyz-services-in-leaflet

#install.packages("proj4")

loc = proj4::project(cas.data[1:2], "+proj=tmerc +lat_0=0 +lon_0=173 +k=0.9996 +x_0=1600000 +y_0=10000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs", inv=TRUE)

```

# Plotting geospatial data

### Plotting crashed data points into open street map tile
```{r}

#install.packages("snippets",,"http://rforge.net/",type="source")
library(snippets)
#install.packages("png")
library(png)

par(mar=rep(0,4))
plot(loc$x, loc$y, pch='.', asp=1/cos(-40.6/180*pi), ty='n')
#overlay with oopen street map
osmap()
points(loc$x, loc$y, pch='.',col = 'red')

```

### Plotting crashes from 2016 to 2019 in Auckland

```{r}

#auckland geo spatial boundaries
q = list(x = c(174.149667174548, 175.383632470626), y = c(-36.554874067828, 
-36.991016113938))


akl = loc$x > range(q$x)[1] & loc$x < range(q$x)[2] & loc$y > range(q$y)[1] & loc$y < range(q$y)[2] & cas.data$crashYear %in% 2016:2019
akl[is.na(akl)] = FALSE

sum(akl)
mean(akl)

#usual residence location points
g = aggregate(df2$Total, list(lon=df2$usual.residence.LONGITUDE,lat=df2$usual.residence.LATITUDE), sum)
plot(0, 0, pch='.', xlim=range(q$x), ylim=range(q$y), asp=1/cos(-40.6/180*pi), ty='n')

#usual work location points
g2 = aggregate(df2$Total, list(lon=df2$usual.work.LONGITUDE,lat=df2$usual.work.LATITUDE), sum)

#plotting it
par(mar=rep(0,4))
plot(0, 0, pch='.', xlim=range(q$x), ylim=range(q$y), asp=1/cos(-40.6/180*pi), ty='n')
snippets::osmap()

#RED - usual work, 
points(g$lon, g$lat,pch=19,cex=sqrt(g$x/max(g$x))*1.2,col="#ff000040")
points(g2$lon, g2$lat,pch=19,cex=sqrt(g2$x/max(g2$x))*1.5,col="#0000ff80")
legend("topright", legend=c("Residence", "Work"),
       col=c("red", "blue"), pch = 16, cex=1, box.col="white")

```

### Using Tessaslation to group centroids of usual residence points

```{r}

ga = g[g$lon > range(q$x)[1] & g$lon < range(q$x)[2] & g$lat > range(q$y)[1] & g$lat < range(q$y)[2], ]
## warning ! bad coordinates - since lat/lon aspect ration is not 1!! jsut a hack ;)

#used the deldir package 
#installed.packages("deldir")
d = deldir::deldir(ga$lon, ga$lat)
plot(0, 0, pch='.', xlim=range(q$x), ylim=range(q$y), asp=1/cos(-40.6/180*pi), ty='n')
snippets::osmap()
points(g$lon, g$lat,pch=19,cex=sqrt(g$x/max(g$x))*1.2,col="#ff000040")
points(g2$lon, g2$lat,pch=19,cex=sqrt(g2$x/max(g2$x))*1.5,col="#0000ff80")
plot(d, add=TRUE, wlines="tess", pch='.')

```

